Cross-Dataset Propensity Estimation for \\Debiasing Recommender Systems

### 2022-09-26 paper/paper.tex
Final submitted version

\subsection{Simulated Environment for Controlling Bias \textcolor{blue}{and Quantization}}
Since selection bias is uncontrollable in a dataset completely drawn from real-world, we have to adopt a simulated environment \cite{krauth2020offline} with both semi-synthetic and synthetic datasets explained below. In our environment, we propose the \textbf{softmax observation model} and introduce a hyperparameter $\beta$ to control the degree of bias. For a rating matrix $R$, the corresponding probability matrix of each rating being observed is $\Pr(R_{u,i}\text{ is observed}) = k\text{softmax}(\beta R_{u,i})$, where $k$ is set so that the expected proportion of observed ratings is controlled. In this observation model, a larger $\beta$ induces more significant bias at an exponential rate. In our experiment we assume constantly $10\%$ of ratings are observed among $1000$ users and $1000$ items.
% \begin{figure}
%   \centering
%   \includegraphics[width=\textwidth]{visualize.png}
%   \caption{Visualizing the effect of controlling bias}
% \end{figure}

\textcolor{blue}{Mention how we control quantization: mapping the same underlying continuous-valued $R_{ui}$ to $n$ different quantities evenly spaced between $0$ and $1$, i.e. 0 and 1 or 0, 0.25, 0.5, 0.75, 1}

@@

\subsection{Results}
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{1.png}
  \caption{The effect of selection bias is not affected by quantization}
\end{figure}
Figure 1 shows the performance on the latent factors dataset of 3 classic algorithms: user-KNN, item-KNN, and SVD matrix factorization. \textcolor{blue}{Performance is evaluated by RMSE on the dense quantized (or not?) rating matrix}. The three differently quantized variants come from the same source distribution (the dataset). Gaussian noise is introduced for simulating real-world user variability. Shaded regions denote the regions between quartiles \textcolor{blue}{overs [number] trials of re-sampled training data}.

@@

\section{Experiment}
We designed experiments to verify the performance of our cross-dataset model. We trained on two datasets (as in section 3) with $\beta =1$ in training set and used three baseline algorithms for comparison. In this paper we assumed the highly quantized dataset is unbiased with $\beta=0$, which is probable when the implicit dataset reflects users' spontaneous choices. Empirically, our framework will work as long as the highly quantized dataset is less biased. We selected the root mean square error (RMSE) and the mean absolute error (MAE) \textcolor{blue}{over the dense rating matrix (?)} as performance metrics.

\subsection{Baselines}
\textbf{Matrix Factorization (MF.)}
As a simple baseline, we adopt the standard matrix factorization that does not adopt propensity estimation or importance weighting \textcolor{blue}{and uses only the training set}.

>>>>

\subsection{Simulated Environment for Controlling Bias and Quantization}
Since selection bias is uncontrollable in a dataset completely drawn from real-world, we have to adopt a simulated environment \cite{krauth2020offline} with both semi-synthetic and synthetic datasets explained below. In our environment, we propose the \textbf{softmax observation model} and introduce a hyperparameter $\beta$ to control the degree of bias. For a rating matrix $R$, the corresponding probability matrix of each rating being observed is $\Pr(R_{u,i}\text{ is observed}) = k\,\text{softmax}(\beta R_{u,i})$, where $k$ is set so that the expected proportion of observed ratings is controlled. In this observation model, a larger $\beta$ induces more significant bias at an exponential rate. In our experiment we assume constantly $10\%$ of ratings are observed among $1000$ users and $1000$ items. Quantization is simulated by mapping the same underlying continuous-valued rating matrix $R$ to $n$ different quantities evenly spaced between $0$ and $1$. In all experiments we used only 2, 3, and 5 as quantization levels.
% \begin{figure}
%   \centering
%   \includegraphics[width=\textwidth]{visualize.png}
%   \caption{Visualizing the effect of controlling bias}
% \end{figure}

@@

\subsection{Results}
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{1.png}
  \caption{The effect of selection bias is not affected by quantization}
\end{figure}
Figure 1 shows the performance on the latent factors dataset of 3 classic algorithms: user-KNN, item-KNN, and SVD matrix factorization. Performance is evaluated by RMSE on the dense ground truth rating matrix. The three differently quantized variants come from the same source distribution (the dataset). Gaussian noise is introduced for simulating real-world user variability. Shaded regions denote the regions between quartiles overs 10 trials of re-sampled training data.

@@

\section{Experiment}
We designed experiments to verify the performance of our cross-dataset model. We trained on two datasets (as in section 3) with $\beta =1$ in training set and used three baseline algorithms for comparison. In this paper we assumed the highly quantized dataset is unbiased with $\beta=0$, which is probable when the implicit dataset reflects users' spontaneous choices. Empirically, our framework will work as long as the highly quantized dataset is less biased. We selected the root mean square error (RMSE) and the mean absolute error (MAE) over the dense ground truth rating matrix as performance metrics.

\subsection{Baselines}
\textbf{Matrix Factorization (MF.)}
As a simple baseline, we adopt the standard matrix factorization that does not adopt propensity estimation or importance weighting and uses only the less quantized training set.
